Notes

### 2022-09-22 notes/notes.tex
restructure and clarify notes

\subsection{Model with time-dependent correction to the emissions}
If we weigh each day and each emission source by some factor, we are applying a transformation
$$
\begin{bmatrix}
    s_1^{-1} x_1 \\
    \vdots \\
    s_1^{-1} x_1 \\ 
    s_2^{-1} x_2 \\
    \vdots \\ 
    s_m^{-1} x_m
\end{bmatrix}
\mapsto
\begin{bmatrix}
    w_{1,1} s_1^{-1} x_1 \\
    \vdots \\
    w_{1,365} s_1^{-1} x_1 \\ 
    w_{2,1} s_2^{-1} x_2 \\
    \vdots \\ 
    w_{m, 365} s_m^{-1} x_m
\end{bmatrix}
$$
which can be written concisely as 
$$
W (S^{-1} x \otimes \vb{1}_{365})
\quad\text{where}\quad 
W = \begin{bmatrix}
    w_{1,1} & 0 & \cdots & 0 \\
    0 & w_{1,2} & \cdots & 0 \\
    \vdots & \vdots & \ddots & 0 \\
    0 & 0 & \cdots & w_{m, 365}
\end{bmatrix}
.$$
Using this weighted vector of emissions instead of $\tilde{x}$ in \cref{eq: initial model 365 days} gives
\begin{equation}
    \label{eq: time-evolving model concise}
\hat{y} = M_{shift} W (S^{-1} x \otimes \vb{1}_{365})
.\end{equation}
In this expression, the unknown is $W$ and should be fit to the data. 
% There are a plethora of numerical techniques for solving linear equations where the unknown is a matrix.
% In the following, we provide a theoretical description of a system that can be solved for $W$.

% We can put constraints on $W$ of the form $W \in \mathcal{W}$, where $\mathcal{W}$ is some set of constrained diagonal matrices. Suppose we have a parametrisation $\Gamma: \mathbb{R}^{k} \rightarrow \mathcal{W}, u \mapsto \Gamma(u)$. 
% In this case, \cref{eq: time-evolving model concise} can be reformulated as a linear operation applied to $\Gamma(u)$ as follows:
% $$
% \hat{y} = ((S^{-1} x \otimes \vb{1}_{365})^T \otimes M_{shift}) \mathrm{vec}\, \Gamma(u)
% .$$
% If $\Gamma$ is a linear map, the above is a linear expression in $u$. Then, the minimisation of $\norm{y - \hat{y}}^2$ can be solved using numerical linear algebra.

\section{Fitting the model}
\subsection{Regularisation}
The indices $i = 1,\dots,n$ are partitioned as a disjoint union $\mathcal{I}_{det} \cup \mathcal{I}_{0}$ where $\mathcal{I}_0$ contains the indices of non-detection events. Let $P_{det}$ be the linear map that removes the rows of a matrix that correspond to $\mathcal{I}_0$. We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise \emph{natural} logarithm. The simplicity of $W$ can be measured as $\norm{\log W}^2$. If we forget about non-detections, we have the cost function 
$$
\min_{W \text{ diagonal}} f(W) =
\underbrace{\frac{1}{2}\norm{\log.(P_{det} M_{shift} W \tilde{x})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
.$$
In order to optimise $f$ numerically, we need to be able to measure changes to $W$. We do this implicitly by parametrising $W = e^{\mathrm{diag}\, v}$ for some vector $v$. If we apply an absolute change $v' := v + \Delta v$ in numerical optimisation, this corresponds to the relative change $W' = W e^{\mathrm{diag} \Delta v}$. 
Reformulating the cost function gives 
$$
\min_{v} f(v) =
\underbrace{\frac{1}{2}\norm{\log.(P_{det} M_{shift} e^{\mathrm{diag}\, v} \tilde{x})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{v}^2}_{\text{deviation from average emission}}
.$$

\begin{proposition}
    Let $\hat{z}$ be the vector $[\hat{z}_1\,\,\dots\,\,\hat{z}_n]^T$ where 
    $$
    \hat{z}_i = \begin{cases}
        \frac{\log \hat{y}_i}{\hat{y}_i} & \text{if observation $i$ is a detection} \\
        0 & \text{if observation $i$ is a non-detection} 
    \end{cases}
    .$$
    Then
    $$
    \mathrm{grad}_{v} f(v) = e^v \odot (M_{shift}^T \hat{z}) \odot \tilde{x} + \lambda v
    .$$
\end{proposition}
\begin{proof}

Let $U = \mathrm{vec} \circ \mathrm{diag}$. Then $\mathrm{vec}\, W(v) = U e^v$ where $e^v$ denotes elementwise exponentiation.
Define $$
G(v) :=
P_{det} \hat{y}_i
=
P_{det} M_{shift} W(v) \tilde{x}
\quad\text{and}\quad
F(v) := \log.(G(v))
.$$
Then 
$$
G(v) = P_{det} M_{shift} W(v) \tilde{x}
=
(\tilde{x}^T \otimes P_{det} M_{shift}) \mathrm{vec} W(v)
=
\underbrace{(\tilde{x}^T \otimes P_{det}M_{shift}) U}_A e^v
$$
so that 
$$
\frac{\partial G}{\partial v} = A e^{\mathrm{diag}\, v}
.$$
Differentiating $F(v) = \log.(G(v))$ gives:
$$
\frac{\partial F}{\partial v}
=
\frac{\partial}{\partial v} \log.(G(v))
=
\mathrm{diag}\,(G(v))^{-1} \frac{\partial G}{\partial v}
=
\mathrm{diag}\,(G(v))^{-1} A e^{\mathrm{diag}\, v }
.$$
Write $z := P_{det} \hat{z}$.
We can interpret $z$ as the vector obtained by applying the map $x \mapsto (\log x) / x$ elementwise to $G(v)$.
Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
\begin{align*}
\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right) =\left(\frac{\partial F}{\partial v}\right)^T F(v)
&= e^{\mathrm{diag}\, v} A^T \mathrm{diag}(G(t))^{-1} \log.(G(t)) \\
&= e^{\mathrm{diag}\, v} A^T z \\
&= e^{\mathrm{diag}\, v} U^T (\tilde{x} \otimes (P_{det} M_{shift})^T) z \\
&= e^{\mathrm{diag}\, v} U^T (P_{det} M_{shift})^T z \tilde{x}^T \\
&= e^{\mathrm{diag}\, v} U^T M_{shift}^T P_{det}^T P_{det} \hat{z} \tilde{x}^T \\
&= e^{\mathrm{diag}\, v} U^T M_{shift}^T \hat{z} \tilde{x}^T
\end{align*}
where the last equality follows from the fact that $P_{det}^T P_{det}$ is the map that sets the rows corresponding to the non-detection events to zero. 
By construction of $U$, the map $U^T$ projects its argument orthogonally onto the diagonal matrices and gives the diagonal element. The orthogonal projection of a matrix onto the diagonal matrices is obtained by setting the off-diagonal elements to zero. The diagonal elements of $M_{shift}^T \hat{z} \tilde{x}^T$ are $(M_{shift}^T \hat{z}) \odot \tilde{x}$, so that the above simplifies to
$$
\mathrm{grad}_v\, \left(\frac{1}{2} \norm{F(v)}^2 \right)
=
e^{\mathrm{diag}\, v} \left((M_{shift}^T \hat{z}) \odot \tilde{x}\right)
=
e^v \odot (M_{shift}^T \hat{z}) \odot \tilde{x}
.$$

The foregoing gives the derivative of the first term in $f(W)$. The gradient of the second term can is simply $\lambda v$.

\end{proof}
For even more evidence, there is a Julia script in the repository that checks this formula against the result obtained from automatic differentiation (using simplified mock data). The result is correct up to machine precision.

\subsection{Constrained optimisation (similar idea as regularisation)}
Constrained optimisation is similar to using regularisation.
Suppose that the number of components in $\tilde{x}$ is $m T$ (where $m$ is the number of sources and $T$ is the number of days).
If we want to bound the average of $(\log w_{ij})^2$ by some value $\delta^2$, then we have the constrained optimisation problem
\begin{align*}
    \min\quad & \frac{1}{2} \norm{\log.(P_{det} M_{shift} W \tilde{x})}^2 \\
    \text{s.t.}\quad & \frac{1}{mT} \norm{\log W}^2 \le \delta^2
\end{align*} 
or equivalently 
\begin{align*}
    \min\quad & \frac{1}{2} \norm{\log.(P_{det} M_{shift} e^{\mathrm{diag}\, v} \tilde{x})}^2 \\
    \text{s.t.}\quad & \frac{1}{mT} \norm{v}^2 \le \delta^2
\end{align*}
which can be solved using very similar ideas as the regularisation section.

\section{Correcting for non-detections}

@@

Now a good cost function is
$$
\min_{v} f(v) =
\underbrace{\frac{1}{2}\norm{\log h.(\hat{y})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{v}^2}_{\text{deviation from average emission}}
\quad\text{where}\quad 
\hat{y} = M_{shift} e^{\mathrm{diag}\, v} \tilde{x}
$$
where $\log h.$ is the elementwise application of $x \mapsto \log h(x)$ to a vector.

\begin{proposition}
    Let $g$ be the vector defined elementwise by 
    $$
    g_i = \frac{\log h(\hat{y}_i)}{h(\hat{y}_i)} \frac{\mathrm{d}}{\mathrm{d}\hat{y}_i} h(\hat{y}_i)
    .$$
    Then 
    $$
    \mathrm{grad}\, f(v) = e^v \odot (M_{shift}^T g) \odot \tilde{x} 
    $$
\end{proposition}
\begin{proof}

@@

\section{Further work}
\begin{itemize}
    \item For certain emission sources in certain periods, there is an expected correlation between the emission at time $t$ and at time $t + 1, t + 2, \dots$. We can add a constraint on the autocorrelation.
    \item Change the parametrisation to something like this: 
    \begin{align*}
        x_1^{2 \text{ Jan } 2014} &= \alpha_1^{(1)} x_1^{1 \text{ Jan } 2014} \\
        x_1^{3 \text{ Jan} 2014} &= \alpha_2^{(1)} x_1^{2 \text{ Jan } 2014} \\
        &\vdots
    \end{align*}
    with a constraint like $\alpha_{\min} \le \prod_{t=1}^T \alpha_t^{(j)} \le \alpha_{\max}$ for all $T$ and all $j = 1,\dots,m$. However, this would be difficult to implement. In addition, the complexity of inequality constrained optimisation can be combinatorial in the number of constraints if an active set method is used \cite{Nocedal2006}.

    \item How do you store the sparse matrix $M_{shift}$ in a computer? Most programming languages have data structures for large sparse matrices?
\end{itemize}

>>>>

\subsection{Model with time-dependent correction to the emissions}
If we weigh each day and each emission source by some factor, we are applying a transformation
$$
\begin{bmatrix}
    s_1^{-1} x_1 \\
    \vdots \\
    s_1^{-1} x_1 \\ 
    s_2^{-1} x_2 \\
    \vdots \\ 
    s_m^{-1} x_m
\end{bmatrix}
\mapsto
\begin{bmatrix}
    w_{1,1} s_1^{-1} x_1 \\
    \vdots \\
    w_{1,365} s_1^{-1} x_1 \\ 
    w_{2,1} s_2^{-1} x_2 \\
    \vdots \\ 
    w_{m, 365} s_m^{-1} x_m
\end{bmatrix}
$$
which can be written concisely as 
$$
W (S^{-1} x \otimes \vb{1}_{365})
\quad\text{where}\quad 
W = \begin{bmatrix}
    w_{1,1} & 0 & \cdots & 0 \\
    0 & w_{1,2} & \cdots & 0 \\
    \vdots & \vdots & \ddots & 0 \\
    0 & 0 & \cdots & w_{m, 365}
\end{bmatrix}
.$$
Using this weighted vector of emissions instead of $\tilde{x}$ in \cref{eq: initial model 365 days} gives
\begin{equation}
    \label{eq: time-evolving model concise}
\hat{y} = M_{shift} W (S^{-1} x \otimes \vb{1}_{365})
.\end{equation}
In this expression, the unknown is $W$ and should be fit to the data.

\section{Correction for non-detections}

@@

\section{Fitting the model}
In the above, we argued that, the prediction is accurate if $h(\hat{y})$ is close to one, where $\hat{y}$ is the predicted concentration. This is equivalent to $\log^2(h(\hat{y}))$ being small. Notice that this penalises overestimates by a factor $\alpha$ as much as underestimates by a factor $\alpha$. That is, 
$$
\log^2(\alpha) = \log^2(\alpha^{-1})
.$$
The average prediction error over all predictions is $\sum_{i=1}^n \log^2(h(\hat{y}_i)) = \norm{\log h(\hat{y})}^2$. This will be the first criterion for optimality of the model.

The second criterion is the ratio between the daily emissions and the average emission for each source, i.e. the factors $W$. In a similar vein as the above, the size of $W$ can be measured as $\norm{\log W}^2$ where $\norm{\cdot}$ is the Frobenius norm\footnote{Since $W$ is a diagonal matrix, the matrix logarithm $\log W$ is simply the diagonal matrix with $\log w_{1,1}, \log w_{1,2}, \dots$ on the diagonal. Hence, $\norm{\log W}^2 = \log^2 w_{1,1} + \log^2 w_{1,2} + \dots$}.

To balance these two criteria, we measure the quality of the model as 
\begin{equation}
    \label{eq: cost function regularised}
f(W) := 
\underbrace{\frac{1}{2}\norm{\log h.(\hat{y})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
\quad\text{where}\quad 
\hat{y} = M_{shift} W \tilde{x}
\end{equation}
where $\lambda \ge 0$ is a regularisation parameter.

To optimise $f$ numerically, we will parametrise $W$ as $W = e^{\mathrm{diag}\, v}$ where $v = [v_1, v_2,\dots]$ is some vector and $e^{\mathrm{diag}\, v}$ is the diagonal matrix with $[e^{v_1}, e^{v_2},\dots]$ on the diagonal. The motivation for doing this is twofold:
\begin{enumerate}
    \item It ensures that $W$ remains strictly positive. In particular, the diagonal elements of $W$ can only tend towards zero if $v_j \rightarrow -\infty$ for some $i$. 
    \item It implicitly measures changes to $W$ in a relative sense. If the optimisation algorithm performs a linear update $v_{k+1} = v_k + \Delta v$, this corresponds to a relative update $W_{k+1} = W_k e^{\mathrm{diag}\, \Delta v}$.
\end{enumerate}

Thus, in practice, we are interested in the optimisation problem
$$
\min_{v} f(v) =
\frac{1}{2}\norm{\log h.(\hat{y})}^2 + \frac{1}{2}\lambda \norm{v}^2
\quad\text{where}\quad 
\hat{y} = M_{shift} e^{\mathrm{diag}\, v} \tilde{x}
.$$
Most optimisation algorithms require at least the gradient of $f$. This is given by the following:
\begin{proposition}
    Let $g = [g_1, \dots, g_n]$ be the vector defined elementwise by 
    $$
    g_i = \frac{\log h(\hat{y}_i)}{h(\hat{y}_i)} \frac{\mathrm{d}}{\mathrm{d}\hat{y}_i} h(\hat{y}_i)
    .$$
    Then 
    $$
    \mathrm{grad}\, f(v) = e^v \odot (M_{shift}^T g) \odot \tilde{x} 
    $$
    where $e^v$ denotes elementwise exponentiation and $\odot$ is the Hadamard (elementwise) product.
\end{proposition}
\begin{proof}

@@

\section{Further work}
\begin{itemize}
    \item For certain emission sources in certain periods, there is an expected correlation between the emission at time $t$ and at time $t + 1, t + 2, \dots$. We can add a constraint on the autocorrelation.
    \item Time-correlation could be measured by parameterising the model as follows: 
    \begin{align*}
        x_1^{2 \text{ Jan } 2014} &= \alpha_1^{(1)} x_1^{1 \text{ Jan } 2014} \\
        x_1^{3 \text{ Jan} 2014} &= \alpha_2^{(1)} x_1^{2 \text{ Jan } 2014} \\
        &\vdots
    \end{align*}
    with a constraint like $\alpha_{\min} \le \prod_{t=1}^T \alpha_t^{(j)} \le \alpha_{\max}$ for all $T$ and all $j = 1,\dots,m$. However, this would be difficult to implement. In addition, the complexity of inequality constrained optimisation can be combinatorial in the number of constraints if an active set method is used \cite{Nocedal2006}.

    \item For the sake of simplicity, the current implementation does not store $M_{shift}$ in memory in a data structure for large and sparse matrices. This could be changed to improve memory use and/or computation time.

    \item Other numerical optimisation algorithms may be more efficient or have better convergence properties. For instance, the Gauss-Newton and Levenberg-Marquardt methods are specifically designed for optimising the squared norm of a vector-valued function, but may require a large amount of memory \cite[Section 10.3]{Nocedal2006}.

    \item Standard software libraries support optimisation with constraints. This is an alternative to regularisation and may be more interpretable.
    Suppose that the number of components in $\tilde{x}$ is $m T$ (where $m$ is the number of sources and $T$ is the number of days).
    If we want to bound the average of $(\log w_{ij})^2$ by some value $\delta^2$, then we have the constrained optimisation problem
    \begin{align*}
        \min\quad & \frac{1}{2} \norm{\log h.(\hat{y})}^2 \\
        \text{s.t.}\quad & \frac{1}{mT} \norm{\log W}^2 \le \delta^2
    \end{align*} 
    which can be solved using similar ideas as those developed above. The prototypical constrained optimisation method (such as an active set method) would iterate towards a critical point of the Lagrangian
    $$
    \mathcal{L}(W, \lambda) = \frac{1}{2} \norm{\log h.(\hat{y})}^2 + \lambda\left(\frac{1}{mT} \norm{\log W}^2 - \delta^2\right)
    $$
    in which case the computational cost would be comparable to optimising \cref{eq: cost function regularised}. A more detailed explanation of constrained optimisation and the Lagrangian is found in \cite{Nocedal2006}.
\end{itemize}
